Ages of the Kepler Stars and Planets

### 2024-04-24 paper/ms.tex
update ms

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have yielded detailed
  pictures of how the rotation rates and lithium content of FGK stars
  evolve during the first few billion years of their lives.
  %
  Here we apply this knowledge to planets and stars observed by the
  primary Kepler mission, and report the resulting rotation and
  lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is probably valid.
  %
  The independent lithium ages show X,Y,Z, though they are subject to their own systematic issues.
  %
  The relative statistical precisions from rotation, $\sigma_{t_{\rm
  gyro}} / t_{\rm gyro}$, are XX +YY -ZZ\%, which is X.X +Y.Y -Z.Z more
  statistically precise than isochronal ages for the same stars, and
  $\approx$X.X times less precise than cluster-based ages.
  %
  The results include \nplyounggyrotwosigma\
  planets younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\
  planets at 3$\sigma$).
  %
  While lithium provides only marginal added information for most of
  this sample,
  we report new robust ages from both rotation and lithium for 14
  confirmed $<$500\,Myr planets, including five mini-Neptunes, five
  super-Earths, three Earths and sub-Earths, and a Neptune-sized
  planet.
  %
  In the broader Kepler target star sample, we find that the
  gyrochrone age distribution rises linearly from 0 to 3\,Gyr, with
  stars 2\,Gyr old being twice as common in the Kepler field as stars
  1\,Gyr old.
  %
  The resulting age distribution bounds the relative balance of
  ``young'' and ``old'' planets detectable not only by Kepler, but
  also by future wide-field transit surveys.
  %
%  The sample of detected Kepler planets has a similar age
%  distribution, though it is slightly deficient at $\lesssim$0.2\,Gyr,
%  likely due to decreased detection sensitivity.
  %
  %Combining our stellar age catalog with Kepler's completeness products, we perform
  %a hierarchical bayesian analysis to infer the average number of planets per star
  %as a function of both stellar and planetary parameters.
  %
  %While we recover many previously known correlations with stellar mass and metallicity,
  %our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
  %and in influencing the evolution of the mini-Neptune and super-Earth populations.
  %TODO will this be the case?
%  We also comment on a few previously reported evolutionary trends in
%  the size evolution of close-in mini-Neptunes (2-4\,\rearth) and
%  super-Earths (1-2\,\rearth), a few of which we recover at comparable
%  ($\sim$3$\sigma$) significance to previous work, and one of which we
%  note for the first time.
\end{abstract}

@@

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions.  Here, we will focus on planets previously found using
transits, and infer new stellar ages using rotation and lithium.
Rotation will provide the most age information for the largest number
of stars.  The lithium technique will provide useful information for
stars younger than a few hundred million years.

@@

	\caption{
		{\bf Sizes, orbital periods, and rotation-based ages of transiting exoplanets younger than one billion years.} 
		%		
		%		Two possible selection functions for what defines a young, sub-gigayear, planet
		%		are imposed on the {\it top} and {\it bottom}: the former requires $t$<1\,Gyr with
		%		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
		%		Circles are previously reported measurements from the literature.
		Stars denote \nplyounggyrotwosigmanograzingnoruwe\ systems that are newly characterized in this work to have
		$t_{\rm gyro}$<1\,Gyr at $2$$\sigma$; circles
		are planets from the literature meeting the same age requirement, but with a heterogeneous
		set of age provenances.
		Gray points are transiting planets from the NASA Exoplanet Archive older than
		1\,Gyr.
		The selection function matters:
		selecting for precise ages ($t/\sigma_{\rm t}>3$) would hide $\approx$ten
		$<$0.5\,Gyr planets, and add $\approx$twenty 0.5-1\,Gyr planets.
		For this plot, we require our Kepler stellar hosts to have no quality flags raised,
		and for the planets to similarly no have quality flags raised. 
		Weakening these constraints would expand the sample at the expense of reliability.
		\label{fig:rp_period_age_results}
	}
\end{figure*}

>>>>

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have shown how the rotation
  rates and lithium content of FGK stars evolve during the first few
  billion years of their lives.
  %
  Here we apply this knowledge to planets and stars observed by the
  primary Kepler mission, and calculate the implied rotation and
  lithium-based ages.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is probably valid.
  %
  The independent lithium ages show X,Y,Z, though they are subject to their own systematic issues.
  %
  The relative statistical precisions from rotation, $\sigma_{t_{\rm
  gyro}} / t_{\rm gyro}$, are XX +YY -ZZ\%, which is X.X +Y.Y -Z.Z more
  statistically precise than isochronal ages for the same stars, and
  $\approx$X.X times less precise than cluster-based ages.
  %
  The results include \nplyounggyrotwosigma\
  planets younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\
  planets at 3$\sigma$).
  %
  While lithium provides only marginal added information for most of
  this sample,
  we report new robust ages from both rotation and lithium for 14
  confirmed $<$500\,Myr planets, including five mini-Neptunes, five
  super-Earths, three Earths and sub-Earths, and a Neptune-sized
  planet.
  %
  In the broader Kepler target star sample, we find that the
  gyrochrone age distribution rises linearly from 0 to 3\,Gyr, with
  stars 2\,Gyr old being twice as common in the Kepler field as stars
  1\,Gyr old.
  %
  This parent stellar age distribution bounds the relative balance of
  young and old planets detectable not only by Kepler, but also by
  future wide-field transit surveys.
  %
%  The sample of detected Kepler planets has a similar age
%  distribution, though it is slightly deficient at $\lesssim$0.2\,Gyr,
%  likely due to decreased detection sensitivity.
  %
  %Combining our stellar age catalog with Kepler's completeness products, we perform
  %a hierarchical bayesian analysis to infer the average number of planets per star
  %as a function of both stellar and planetary parameters.
  %
  %While we recover many previously known correlations with stellar mass and metallicity,
  %our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
  %and in influencing the evolution of the mini-Neptune and super-Earth populations.
  %TODO will this be the case?
%  We also comment on a few previously reported evolutionary trends in
%  the size evolution of close-in mini-Neptunes (2-4\,\rearth) and
%  super-Earths (1-2\,\rearth), a few of which we recover at comparable
%  ($\sim$3$\sigma$) significance to previous work, and one of which we
%  note for the first time.
\end{abstract}

@@

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions.  Here, we will focus on planets previously found using
transits, and infer new stellar ages using rotation and lithium.
Rotation will provide the most age information for the largest number
of stars.  Lithium will provide useful information for
stars younger than a few hundred million years.

@@

	\caption{
    {\bf Sizes, orbital periods, and rotation-based ages of transiting
    exoplanets younger than one billion years.} Stars denote
    \nplyounggyrotwosigmanograzingnoruwe\ planets that are newly
    characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
    $2$$\sigma$; circles are planets from the literature meeting the
    same age requirement, but with a heterogeneous set of age
    provenances.  Gray points are transiting planets from the NASA
    Exoplanet Archive older than 1\,Gyr.  The selection function
    matters: selecting for precise ages ($t/\sigma_{\rm t}>3$) would
    hide $\approx$ten $<$0.5\,Gyr planets, and add $\approx$twenty
    0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
    hosts to have no quality flags raised, and for the planets to
    similarly no have quality flags raised.  Weakening these
    constraints would expand the sample at the expense of reliability.
		\label{fig:rp_period_age_results}
	}
\end{figure*}
